---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
options(width = 120)

```

Estimate and test mixed-effects signal detection theory models with maximum likelihood estimation. 
This package leverages the equivalence between a subclass of SDT and a subclass of generalized linear models first shown by DeCarlo (1998) to estimate mixed SDT models using 
software for generalized linear mixed models, that is, the `lme4` or `glmmTMB` 
package. Mixed-effects SDT models can be specified on the level of SDT 
parameters discriminability and response bias using the typical R formula syntax:

` discriminability = ~ X + (X | ID)`

` bias = ~ X + (X | ID)` 

`mesdt` translates the given model formulas to a GLMM, uses either `lme4` or `glmmTMB` to estimate the model (based on the user-specified backend), and transforms the parameters back to SDT space, such that all post-processing (e.g., estimating marginal means) can take place on the level of SDT parameters as well. Hypothesis tests related to fixed and random effects can be conducted with Wald tests (only for fixed effects), likelihood ratio tests (type II and type III), and tests based on parametric bootstrapping (type II and type III). 

## Getting Started

### Step 0: Installation & Setup

You can install the development version of `mesdt` from [GitHub](https://github.com/) with:

``` {r, eval = FALSE}
# install.packages("devtools")
devtools::install_github("marie-jakob/mesdt")
```

Since `mesdt` uses `lme4` as the default backend, you also need a working 
installation of `lme4` and, if you want to use it, `glmmTMB`. 

``` {r}
library(mesdt)
```

Upon loading the package, it tells us that it has set the backend, that is, the
package used to estimate the GLMMs to `lme4`. If `glmmTMB` (which can be 
significantly faster) is installed, the backend can be changed like so:

``` {r, eval = FALSE}
options("mesdt.backend" = "glmmTMB")
```

### Step 1: Specify and Fit a Mixed-Effects SDT Model

We are using the `debi3` dataset provided with this package, where we 
investigated attributions to gender discrimination using a signal detection 
approach. In the experiment, male and female participants 
(as indicated by the variable `participant gender`) had to judge 256 
fictional pay raise decisions as biased or unbiased. The cases involved male 
and female employees(`emp_gender`), who were either granted or denied a pay raise 
(`committee`). 

```{r}
debi3 <- mesdt::debi3

```

To estimate effects of these variables on participants' discriminability and 
response bias, we include them as fixed effects in our model. To account for
differences between our participants in these parameters, we include 
by-participant random intercepts and slopes on discriminability and response bias. 
Our studies with this paradigm consistently showed considerable by-participant
variability in mean response bias, mean discriminability and the effect of the 
committee decision on participants' response bias, which is why we include
by-participant random intercepts on both SDT parameters and a by-participant
random slope on response bias.

`mesdt` also requires the user to specify the variable indicating the type of
trial (i.e., whether the given trial was a signal, coded as 1, or noise trial, 
coded as 0, `status` in `debi3`) and the variable indicating participants' 
response (`assessment` in `debi3`). 

Thus, in `mesdt`, we can specify and fit our model like this:

``` {r fit}

mod <- fit_mesdt(
  discriminability ~ committee * emp_gender * participant_gender + (1 | id),
  bias ~ committee * emp_gender * participant_gender + (committee | id),
  data = debi3,
  trial_type_var = "status",
  dv = "assessment"
)

summary(mod)

```

The `summary()` method prints population-level estimates for the fixed effects 
we specified separate for discriminability and response bias in a similar format
as`lme4` and `glmmTMB`. The p values in the last column are results from Wald
tests based on the beta estimates and their standard errors. 

### Step 2: Compute Hypothesis Tests for Selected Parameters

Since Wald tests are often not recommended for inference in GLMM, `mesdt` allows
the user to compute different types (see the documentation for details) 
of likelihood ratio tests and parametric bootstrapping tests, that are based on 
comparisons of nested models. 

`mesdt` allows the user to either specify the to-be-tested parameters via 
formulas (e.g., `~ committee`) or to test all fixed effects (`"all"`). For this
experiment, we were interested (among other things) if participants' response 
bias and / or discriminability vary as a function of the committee decision:


```{r test}
fit <- fit_mesdt(~ x1 + (x1 | ID), ~ x1 + (x1 | ID), dv = "y", data = internal_sdt_data,
                 trial_type_var = "trial_type_fac")

lrts_test <- compute_tests(fit, test_intercepts = T)

tests <- compute_tests(mod, 
                       tests = "lrt",
                       tests_discriminability = ~ committee,
                       tests_response_bias = ~ committee)

tests$LRTs$LRT_results

```

In this subset of our data, there is a significant effect of the committee 
decision on discriminability, but not on response bias. 


### Step 3: Estimate marginal means

`mesdt` additionally provides a custom `emmeans()` function, allowing the user
to estimate marginal means for response bias and discriminability. The syntax
is the typical `emmeans` syntax with an additional argument `dpar` specifying 
the SDT parameter. 
Thus, we can get estimated marginal means for response bias and discriminability
for "denied" and "granted" decisions like so: 

```{r emmeans}
library(emmeans)

emmeans(mod, ~ committee, dpar = "discriminability")
emmeans(mod, ~ committee, dpar = "response bias")

```

The estimated marginal means show that participants' discriminability was higher 
in trials where a pay raise was denied than for cases where one was granted. 
Descriptively,  


## References

TODO
